  
---
output: rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, comment="#>")
library(magrittr)
library(ggplot2)
library(ggpubr)
```

# PPI-Context

Contextualization of protein-protein interaction databases by cell line

#### Clone repository
```
$ git clone https://github.com/montilab/ppi-context
```

#### Install requirements
```
$ cd ppi-context
$ pip install -r requirements.txt
```

#### The data

If you just want the data it's easy to load into R...

```
$ R
```

```{r}
ppi <- read.delim("data/v_1_00/PPI-Context.txt", header=TRUE, sep="\t", stringsAsFactors=FALSE)
```

```{r, fig.width=9, fig.align='center'}
data.frame(sort(table(ppi$cell_name), decreasing=TRUE)) %>%
set_colnames(c("var", "freq")) %>%
head(30) %>%
ggbarplot(x="var", y="freq", fill="freq") +
labs(title="", x="Cell Line Name", y="PPI") +
scale_fill_viridis_c(option="inferno", begin=0, end=0.8) + 
theme(legend.position="none",
      axis.text.x=element_text(angle=45, hjust=1, size=12, face="bold"))
```

#### Pre-processing the data
```
| PPI - Context (v1.0)
usage: ppictx.py [-h] [-r] [-d]
                   [-fh PATH_HIPPIE] 
                   [-fp PATH_PUBTATOR]
                   [-fc PATH_CELLOSAURUS]

optional arguments:
  -h, --help            show this help message and exit
  -r, --run             run pipeline
  -d, --download        download raw data first
  -fh PATH_HIPPIE       path to downloaded Hippie data (optional)
  -fp PATH_PUBTATOR     path to downloaded Pubtator data (optional)
  -fc PATH_CELLOSAURUS  path to downloaded Cellosaurus data (optional)
```

In most cases you will need to download the latest bulk data first and then process it...
```bash
$ python ppictx.py --download --run
```

```
| PPI - Context (v1.0)
| Downloading raw data...
| Processing raw data
  ~ [PPI]
  ~ [PID -> CLA]
  ~ [CLA -> CID]
  ~ [PPI -> PID -> CLA -> CID]
```

In other cases, you might have the previous versions of the data to process...
```bash
$ python ppictx.py --run \
                   -fh path/to/HIPPIE.mitab \
                   -fp path/to/PUBTATOR.gz \
                   -fc path/to/CELLOSAURUS.txt
```