---
title: "Tissue Specificity"
vignette: >
  %\VignetteIndexEntry{Tissue Specificity}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="", cache=FALSE, fig.align="center", fig.width=12, fig.height=7)
devtools::load_all(".")
repo <- file.path(Sys.getenv("github"), "ppi-context")
source(file.path(repo, "dev/data.R"))
ppi <- data.ppi(file.path(repo, "data/v_1_00/PPI-Context.txt"))
tissues <- read.delim(file.path(repo, "data/v_1_00/Tissues.txt"))

library(magrittr)
library(dplyr)
library(ggplot2)
library(janitor)
library(ComplexHeatmap)
library(circlize)
```

## Requirements

```{r, eval=FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(janitor)
library(ComplexHeatmap)
library(circlize)
```

## Data

Previous efforts to infer environmental-specificity of PPIs includes integration of tissue-specific gene expression from GTEx. Integrating expression information to infer context may be useful, particulary when used with literature-derived annotations. Here we provide quick access to the processed data for expression of `r nrow(tissues)` proteins across `r ncol(tissues)` derived from the Human Integrated Protein-Protein Interaction rEference (HIPPIE) and provide some useful examples for manipulating the data.

```{r, eval=FALSE}
tissues <- read.delim("data/v_1_00/Tissues.txt")
```

```{r}
tissues <- janitor::clean_names(tissues)
```

If a protein is expressed above a certain threshold in a given tissue, it is deemed active (1) in that tissue.

```{r}
tissues[1:5,1:5]
dim(tissues)
```

## Integrating with interactions

Load in the PPIs...

```{r, eval=FALSE}
ppi <- data.ppi("data/v_1_00/PPI-Context.txt")
```

Here we are just taking the top cell lines with the most PPIs to assign tissue specificity.

```{r}
cells <- data.frame(sort(table(ppi$cell_name), decreasing=TRUE)) %>%
    magrittr::set_colnames(c("var", "freq")) %>%
    head(25) %$%
    var %>%
    as.character()

head(cells)
```

```{r}
dfs <- mapply(function(cell) {
    df <- ppi %>%
          dplyr::filter(cell_name == cell) %>%
          dplyr::select(gene_a, gene_b)
}, cells, SIMPLIFY=FALSE)
```

An interaction is active in a tissue if both interacting partners are expressed in that tissue. Here is an example of using this data to create a *PPI x Tissue* matrix of tissue-specificity for the subset of PPIs annotated with the HeLa cell line.

```{r}
df <- dfs[["HeLa"]]
df %<>% dplyr::mutate(ppi = paste(gene_a, gene_b, sep=" / "))
example <- sapply(tissues, function(tissue) {
    tissue.expressed <- rownames(tissues)[tissue == 1]
    df$gene_a %in% tissue.expressed & df$gene_b %in% tissue.expressed
}) %>%
magrittr::set_rownames(df$ppi)
dim(example)
example[1:5,1:5]
```

## Tissue-activity by cell line

The dimensionality of the data is large, rather we can try to summarize it by looking at the total activity of cell-specific PPIs across different GTEx tissues.

````{r}
data <- lapply(dfs, function(df) {
    apply(tissues, 2, function(tissue) {
        expressed <- as.integer(tissue[df$gene_a] == 1 & tissue[df$gene_b] == 1)
        return( sum(expressed) / length(expressed) ) # Normalize by size
    })
})
```

```{r}
mat <- do.call(cbind, data)
mat[1:5,1:5]
```

```{r}
ht <- HeatmapAnnotation(Size=anno_barplot(apply(tissues, 2, sum), gp=gpar(fill="azure2", color="azure4")))
hr <- rowAnnotation(Size=anno_barplot(sapply(dfs, nrow), gp=gpar(fill="azure2", color="azure4")))

mat.scaled <- t(apply(mat, 2, scale))
colnames(mat.scaled) <- rownames(mat)
Heatmap(mat.scaled,
        name="Activity", 
        col=colorRamp2(c(-2, 0, 2), c("cornflowerblue", "white", "red")),
        rect_gp=gpar(col="white"),
        cluster_rows=TRUE,
        cluster_columns=TRUE,
        top_annotation=ht,
        right_annotation=hr,
        clustering_distance_rows="euclidean",
        clustering_method_rows="ward.D",    
        clustering_distance_columns="euclidean",
        clustering_method_columns="ward.D",
        column_title="GTEx Tissues",
        column_names_gp=gpar(fontsize=8, fontface=2),
        column_names_rot=50,
        row_title="Cell-specific PPIs",
        row_names_gp=gpar(fontsize=8),
        show_column_names=TRUE,
        show_row_names=TRUE)
```

Here is a heatmap visualizing overall activity of cell-specific PPIs across GTEx tissues, where each value represents the fraction of cell-specific PPIs active in a tissue. Barplots in the margins show the number of proteins defined as active in each tissue and the total number of PPIs in each cell line.
